---
title: "HW2"
format: html
editor: visual
---

## Preamble

```{r}
#Library Loading
library(sf)
library(here)
library(tidyverse)
library(tmap)
library(viridis)

#data loading
ejscreen <- st_read(here("data", "ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
HOLC <- st_read(here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))
BDobs <- st_read(here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))
```

# Part 1: Legacy of redlining in current environmental (in)justice

#### Create a map of historical redlining neighborhoods

-   neighborhoods colored by HOLC grade
-   an appropriate base map
-   a clear title and legend

```{r}
tm_shape(HOLC) +
  tm_polygons(
    fill = "grade",
    fill.legend = tm_legend(title = "Grade of neighborhoods in LA")
  ) +
    tm_borders(col = "black") +
  tm_title_out("Redlining in Los Angeles")
```
#### Create a table summarizing
- The percentage of census block groups that fall within each HOLC grade
- Also include the percent of census block groups that don’t fall within a HOLC grade
```{r}
percenttable <- HOLC |>
  st_drop_geometry() |>
  dplyr::count(grade, name = "n_blocks") |>
  dplyr::mutate(pct_blocks = 100 * n_blocks / sum(n_blocks))
view(percenttable)
```
#### Create at least two visualizations summarizing current conditions (from the EJScreen data) within HOLC grades using the mean of the following variables (you may combine variables or create separate plots):
- % low income
- Percentile for Particulate Matter 2.5
- percentile for low life expectancy
- Use ggplot for your visualizations! You will first need to calculate mean of each variable grouped by HOLC grade.

