---
title: "HW2"
format: html
editor: visual
---

## Preamble

```{r}
#Library Loading
library(sf)
library(here)
library(tidyverse)
library(tmap)
library(viridis)

#data loading
ejscreen <- st_read(here("data", "ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
HOLC <- st_read(here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))
BDobs <- st_read(here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))
```

# Part 1: Legacy of redlining in current environmental (in)justice

#### Create a map of historical redlining neighborhoods

-   neighborhoods colored by HOLC grade
-   an appropriate base map
-   a clear title and legend

```{r}
tm_shape(HOLC) +
  tm_polygons(
    fill = "grade",
    fill.legend = tm_legend(title = "Grade of neighborhoods in LA")
  ) +
    tm_borders(col = "black") +
  tm_title_out("Redlining in Los Angeles")
```

#### Create a table summarizing

-   The percentage of census block groups that fall within each HOLC grade
-   Also include the percent of census block groups that don’t fall within a HOLC grade

```{r}
percenttable <- HOLC |>
  st_drop_geometry() |>
  dplyr::count(grade, name = "n_blocks") |>
  dplyr::mutate(pct_blocks = 100 * n_blocks / sum(n_blocks))
view(percenttable)
```

#### Create at least two visualizations summarizing current conditions (from the EJScreen data) within HOLC grades using the mean of the following variables (you may combine variables or create separate plots):

-   % low income
-   Percentile for Particulate Matter 2.5
-   percentile for low life expectancy
-   Use ggplot for your visualizations! You will first need to calculate mean of each variable grouped by HOLC grade.

####Plot 1

```{r}
EJLA <- ejscreen %>% filter(CNTY_NAME == 'Los Angeles County') #filtering for just Los Angeles
if(st_crs(HOLC) != st_crs(EJLA)){
  warning("Make sure to match CRS!")
  HOLC <- st_transform(EJLA, crs = st_crs(EJLA))
}

EJLA_HOLC <- st_join(EJLA,HOLC)
view(EJLA_HOLC)
#HOLC <- st_make_valid(HOLC) #running this, because when I tried to marry the data sets, (st_join) I was getting errors, this should eliminate geometry issues, this is what the internet said. 

#making them friends (joining)
#EJLA_HOLC <- st_join(EJLA, HOLC, join = st_intersects)

#Visualization code
tm_shape(HOLC) + 
  tm_polygons(fill = EJLA$LOWINCPCT
              fill = grade) +
  

```

####Plot 2

```{r}
#Visualization code
ggplot(means_by_grade, aes(x = grade, y = mean_pm25_percentile)) +
  geom_col() +
  labs(
    title = "Percentile of 2.5 Particulate Matter by Grade (LA County)",
    x = "Grade",
    y = "2.5 Particulate Matter 2.5 Percentile (block-group mean within grade)", 
  ) +
  theme_minimal(base_size = 12)
```

```{r}
ggplot(means_by_grade, aes(x = grade, y = mean_life_expectancy)) +
  geom_col() +
  labs(
    title = "Life expectancy by Grade (LA County)",
    x = "Grade",
    y = "Mean Life Expectancy (block-group mean within grade)"
  ) +
  theme_minimal(base_size = 12)
```

#### Write a brief paragraph reflecting on these results

# Part 2: Legacy of redlining in biodiversity observations
